---
title: Exploratory Data Analysis Project
author: Larry Henry Jr.
date: June 8, 2021
output: 
  html_document:
    theme: darkly
    highlight: textmate
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

# Introduction

This project is part of the Udacity's Data Analyst Nanodegree Program. I conducted an Exploratory Data Analysis (EDA) on one of the curated data sets, provided by Udacity, from [Prosper](https://www.prosper.com/plp/about/), which is America’s first [marketplace lending platform](https://www.prosper.com/personal-loans), with over $12 billion in funded loans as of June 2021.

This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan status, borrower income, borrower employment status, borrower credit history, and the latest payment information. The dataset contains loans created between 2005-Q4 and 2014-Q1 with a last updated date of 03/11/2014. Here's a link to the [variable definitions](https://docs.google.com/spreadsheets/d/1gDyi_L4UvIrLTEC6Wri5nbaMmkGmLQBk-Yx3z0XDEtI/edit#gid=0) for this dataset.

The main purpose of this project is to summarize the characteristics of variables that can affect the loan status and to get some ideas about the relationships among multiple variables using summary statistics and data visualizations.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(GGally)
library(dplyr)
# install.packages('ggpubr')
library(ggpubr)
library(RColorBrewer)
library(knitr)
# install.packages("kableExtra")
library(kableExtra)
library(gridExtra)
library(tidyr)
library(memisc)
library(scales)
library(bitops)
library(RCurl)
# install.packages('psych')
library(psych)
library('corrplot')
library(reshape2)
```

```{r  echo=FALSE, message=FALSE, warning=FALSE, Global_seetings}
# To set global options that apply to every chunk in your file, call knitr::opts_chunk$set in a code chunk. Knitr will treat each option that you pass to knitr::opts_chunk$set as a global default that can be overwritten in individual chunk headers.
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

knitr::opts_chunk$set(fig.width = 12,
                      fig.height = 8,
                      fig.path = 'Figs/',
                      fig.align = 'center',
                      tidy = TRUE,
                      echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

# chunk option dev="svg" produces very large vector graphics files
knitr::opts_chunk$set(dev = "svg")

# chunk option dev="png" is the default raster graphics format for HTML output
knitr::opts_chunk$set(dev = "png")

# Global option for table formats
options(knitr.table.format = "simple")
```

## Initial Dataset Exploration

First, the following are the object names of the dataset.

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data_Display_headers}

# Load the Data
loandata <-
  read.csv(
    '~/My Projects/Udacity/udacity-projects/Project Explore and Summarize Data/csv/prosperLoanData.csv'
  )

loandata %>%
  names()

```

> Prosper Loan Data contains 113,937 observations and 81 variables.

Next, the following is the initial dataframe's structure.

```{r echo=FALSE, message=FALSE, warning=FALSE, out.width=20, data_frame_structure}
str(loandata)
```

Next, I will use a subset of this dataset for performing an analysis. The main purpose of this analysis is to explore the various factors that affect borrowers' rates. The following are the variables I would like to use.

```{r echo=FALSE, message=FALSE, warning=FALSE, Subseting}
# Subset the dataframe
loandata <- subset(loandata,
                   select = c(LoanNumber, LoanOriginalAmount, LoanOriginationDate, LoanOriginationQuarter,
                              Term, LoanStatus, BorrowerRate, LenderYield, ProsperRating..Alpha., 
                              ProsperScore, ListingCategory..numeric., EmploymentStatus, IsBorrowerHomeowner,
                              CreditScoreRangeLower, CreditScoreRangeUpper, InquiriesLast6Months,
                              PublicRecordsLast10Years, IncomeRange, IncomeVerifiable))

# Listing the column names for the subset data
names(loandata)
```

Based on the variables in this dataset, I will transform **Term** and **ListingCategory** columns from 'numeric' to 'factor'. Then, I will transform the **ListingCreationDate** column from 'factor' to 'Date'. The following is the dataset's structure transformed.

```{r echo=FALSE, message=FALSE, warning=FALSE, Transforming_Data}
# Factorizing rating, category, and term as these are the key variables we'd investigate throughout the dataset

# Term: numeric to factors
loandata$Term <- factor(loandata$Term)
levels(loandata$Term) <- c("1 year", "3 years", "5 years")

# ListingCategory: numeric to factors
names(loandata)[names(loandata) == 'ListingCategory..numeric.'] <- 'ListingCategory' 

loandata$ListingCategory <- factor(loandata$ListingCategory)

levels(loandata$ListingCategory) <- c("NA", "Debt Consolidation",
                                      "Home Improvement", "Business",
                                      "Personal Loan", "Student Use",
                                      "Auto", "Other", "Baby&Adoption",
                                      "Boat", "Cosmetic Procedure",
                                      "Engagement Ring", "Green loans",
                                      "Household Expenses", "Large Purchase",
                                      "Medical/Dental", "Motocycle", "RV",
                                      "Taxes", "Vacation", "Wedding Loans")

# Levels starts from "AA", all the way to "HR" and ""
# Rename ProsperRating column
names(loandata)[names(loandata) == 'ProsperRating..Alpha.'] <- 'ProsperRating'

loandata$ProsperRating = factor(loandata$ProsperRating,
                                levels = c("AA", "A", "B", "C", "D", "E", "HR", "NA"))

loandata$ProsperScore = factor(loandata$ProsperScore)

# Display the structure of the subset data
str(loandata)
```

------

# Univariate Plots Section

In this section, I will perform some preliminary exploration of the transformed dataset. I will run some summaries of the data and create univariate plots to understand the structure of the individual variables in this dataset.

### Categorical Variables

First, let's take a look on various categorical variables starting with Term.
```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, Term}

# Bar chart: Term
ggplot(data = loandata, aes(x = Term)) +
  geom_bar(color = 'black',
           fill = '#96ADBD',
           width = 0.7) +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    vjust = -0.2,
    colour = "black",
    size = 3.5) +
  ggtitle('Number of Loans by Term') +
  scale_y_continuous(name = 'Loan Count',
                     limits = c(0, 90000),
                     breaks = breaks_width(20000)) +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.6))


# Proportion table: Term
loandata$Term %>%
  table() %>%
  prop.table()
```
> Above 90% of the loans in this dataset are with term more than 1 year. Most of them is of 3 years.    

```{r echo=FALSE, message=FALSE, fig.height=8, warning=FALSE, Loan_by_Year}
# set Loan Orignal Date in Year-Month-Date format
loandata$LoanOriginationDate <-
  as.POSIXct(loandata$LoanOriginationDate,
             format = "%Y-%m-%d")
loandata$LoanOriginationDate.year <-
  loandata$LoanOriginationDate %>%
  format('%Y') %>% strtoi()

ggplot(loandata, aes(as.factor(LoanOriginationDate.year))) +
  geom_bar(color = 'black', fill = '#96ADBD', width = 1) +
  ggtitle('Number of Loans by Year') +
  xlab('Year') +
  ylab('Loan Count') +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    vjust = -0.2,
    colour = "black",
    size = 3.5) +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5)) 

summary(loandata$LoanOriginationDate.year)
```

> The chart took a plunge in 2009 and exploded the year after peaking in 2013.

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, LoanStatus}
# Horizontal bar chart:  LoanStatus
ggplot(data = loandata, aes(x = LoanStatus)) +
  geom_bar(color = 'black', fill = '#96ADBD') +
  coord_flip() +
  labs(x = ' ', y = 'Loan Count') +
  ylim(0, 60000) +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    hjust = -0.2,
    colour = "black",
    size = 3) +
  ggtitle('Number of Loans by Loan Status') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Proportion table: LoanStatus
loandata$LoanStatus %>%
  table() %>%
  prop.table()
```
> There are four major loan status: "Defaulted", "Current", "Completed", and "Chargedoff."    

```{r echo=FALSE, message=FALSE, warning=FALSE, Prosper_Rating}
# Bar chart: ProsperRating
ggplot(data = subset(loandata, ProsperRating != ""), aes(x = ProsperRating)) +
  geom_bar(color = 'black', fill = '#96ADBD') +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    vjust = -0.2,
    colour = "black",
    size = 3.5) +
  ggtitle('Number of Loans by Prosper Rating') +
  labs(caption = "*NA excluded") +  
  scale_y_continuous(name = 'Loan Count',
                     limits = c(0, 20000),
                     breaks = breaks_width(5000)) +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Proportion table: ProsperRating
loandata$ProsperRating %>%
  table() %>%
  prop.table()
```
> The above bar chart shows Prosper's rating distribution. The unrated loans are removed in this plot.  

```{r echo=FALSE, message=FALSE, warning=FALSE, Employment_Status}
# Horizontal bar chart: Employment_Status
ggplot(
  data = loandata, aes(x = EmploymentStatus)) +
  geom_bar(color = 'black', fill = '#96ADBD') +
  coord_flip() +
  labs(x = ' ', y = 'Loan Count') +
  ylim(0, 70000) +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    hjust = -0.2,
    colour = "black",
    size = 3) +
  ggtitle('Number of Loans by Employment Status') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Proportion table : Employment_Status
loandata$EmploymentStatus %>%
  table() %>%
  prop.table()
```
> The majority of Prosper users are either employed or have a full-time job.    

```{r echo=FALSE, message=FALSE, warning=FALSE, Homeowner}
# Pie chart: IsBorrowerHomeowner
ggplot(
  data = loandata, aes(x = '', fill = IsBorrowerHomeowner)) +
  geom_bar() +
  coord_polar(theta = 'y') +
  scale_fill_manual(values=c('#96ADBD', '#425E72'))+
  labs(x = NULL, y = NULL) +
  ggtitle('Number of Borrower Homeowners') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

> About half of the users are homeowners.    

```{r echo=FALSE, message=FALSE, warning=FALSE, IncomeRange}
# Horizontal bar chart: IncomeRange
ggplot(
  data = loandata, aes(x = IncomeRange)) + 
  geom_bar(color = 'black', fill = '#96ADBD') +
  coord_flip() +
  labs(x = ' ', y = 'Loan Count') +
  ylim(0, 35000) +
  geom_text(
    aes(label = ..count..),
    stat = "count",
    hjust = -0.2,
    colour = "black",
    size = 3) +
  ggtitle('Number of Loans by Income Range') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Proporation table: IncomeRange
loandata$IncomeRange %>%
  table() %>%
  prop.table()
```
> About 70% of borrowers' income are in the range of $25,000 - $100,000    

```{r echo=FALSE, message=FALSE, warning=FALSE, IncomeVerifiable}
# Pie chart: IncomeVerifiable
ggplot(
  data = loandata, aes(x = '', fill = IncomeVerifiable)) +
  geom_bar() + 
  coord_polar(theta = 'y') +
  scale_fill_manual(values=c('#96ADBD', '#425E72')) +
  labs(x = NULL, y = NULL) +
  ggtitle('Number of Income Verified Borrowers') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

> Most of borrowers have their income verified.    

### Numerical Variables

Next, let's take a look on various numerical variables starting with BorrowerRate.

```{r echo=FALSE, message=FALSE, BorrowerRate_LenderYield}
# Histogram: BorrowerRate
ggplot(
  data = loandata, aes(x=BorrowerRate)) +
  geom_histogram(color = 'black', fill = '#96ADBD') +
  ggtitle('Number of Loans by Borrower Rate') +
  ylab('Loan Count') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Summary: BorrowerRate
summary(loandata$BorrowerRate)

# Histogram: LenderYield
ggplot(
  data = loandata, aes(x=LenderYield)) + 
  geom_histogram(color = 'black', fill = '#96ADBD') +
  ggtitle('Number of Loans by Lender Yield') +  
  ylab('Loan Count') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Summary: LenderYield
summary(loandata$LenderYield)
```
> The BorrowerRate has a right-skewed distribution, with a mean of 19.28%, and median of 18.40%. Similarly, the LenderYield also has a right-skewed distribution, with a mean of 18.27%, and median of 17.30%.

```{r echo=FALSE, fig.height=8, message=FALSE, warning=FALSE, CreditScore}
# Histogram (p1) : CreditScoreRangeLower
p1 <- ggplot( data = loandata, aes(x=CreditScoreRangeLower)) +
  geom_histogram(binwidth = 10, color = 'black', fill = '#96ADBD') +
  scale_x_continuous(limits = c(400,900)) +
  ylab('Loan Count')+
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

# Histogram (p2) : CreditScoreRangeUpper
p2 <- ggplot(data = loandata, aes(x=CreditScoreRangeUpper)) +
  geom_histogram(binwidth = 10, color = 'black', fill = '#96ADBD') +
  scale_x_continuous(limits = c(400,900)) +
  ylab('Loan Count')+
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))  

# Plot p1 and p2 in the same column
grid.arrange(
  p1,
  p2,
  ncol = 1,
  top = text_grob(label = 'Number of Loans by Credit Score Range',
                  just = 'right'))
```

> The ranges of credit scores are small compared to the distributions. I will use lower credit score for further analyses. 

```{r echo=FALSE, message=FALSE, warning=FALSE, InquiriesLast6Months}
# Histogram: InquiriesLast6Months
ggplot(data = subset(loandata, is.na(loandata$InquiriesLast6Months)!=TRUE),
       aes(x=InquiriesLast6Months)) +
  geom_histogram(binwidth = 1, color = 'black', fill = '#96ADBD') +
  ggtitle('Number of Loans by Inquiries Last 6 Months') +  
  ylab('Loan Count') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))  

# Check unique values
unique(loandata$InquiriesLast6Months)
```

> The distribution of InquiriesLast6Months is right-skewed and long-tailed. Outliers like 105 and 53 exist.

```{r echo=FALSE, message=FALSE, warning=FALSE, PublicRecordsLast10Years}
# Histogram: PublicRecordsLast10Years
ggplot(data = subset(loandata, is.na(loandata$PublicRecordsLast10Years) != TRUE),
       aes(x = PublicRecordsLast10Years)) +
  geom_histogram(binwidth = 1,
                 color = 'black',
                 fill = '#96ADBD') +
  ggtitle('Number of Loans by Public Records Last 10 Years') +  
  ylab('Loan Count') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))   

# Check unique values
unique(loandata$PublicRecordsLast10Years)
```
> The distribution of PublicRecordsLast10Years is right-skewed and has some outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE, LoanOriginalAmount }
# Histogram: LoanOriginalAmount
ggplot(data = loandata, aes(x = LoanOriginalAmount)) +
  geom_histogram(color = 'black', fill = '#96ADBD') +
  ggtitle('Number of Loans by Loan Original Amount') +  
  ylab('Loan Count') +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))    

ggplot(data = loandata, aes(x = LoanOriginalAmount)) +
  geom_histogram(binwidth = 2000,
                 color = 'black',
                 fill = '#96ADBD') +
  labs(title = 'Number of Loans by Loan Original Amount', caption = '*Produced using a binwidth adjusted by 2000 and quantile function') +
  ylab('Loan Count') +
  scale_x_continuous(limits = c(0, quantile(loandata$LoanOriginalAmount, 0.99,
                                            na.rm = TRUE)),
                     breaks = seq(0, quantile(loandata$LoanOriginalAmount, 0.99,
                                na.rm = TRUE), 5000)) +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5))

summary(loandata$LoanOriginalAmount)
```

> LoanOriginalAmount also displays a long-tailed distribution, with some numbers especially higher. The reason why these numbers are more frequent might be because they are the exact numbers (eg. 10,000, 15,000, 20,000) that people tend to choose with a loan original amount minimum of $1,000 and maximum of $35,000

### Univariate Analysis

#### What is the structure of your dataset?

This is a very large dataset with 113937 records and 81 variables on each loan and covers the period from 2005 through 2014. For simplicity with the analysis, I use a subset of the original dataset containing the same number of records, but only with 15 variables that I want to explore. The 15 variables are as follows:

**Categorical:** 
    
    * Term
    * LoanStatus
    * ProsperRating
    * ListingCategory
    * EmploymentStatus
    * IsBorrowerHomeowner
    * IncomeRange
    * IncomeVerifiable

**Numerical:** 

    * BorrowerRate
    * LenderYield
    * CreditScoreRangeLower
    * CreditScoreRangeUpper
    * InquiriesLast6Months
    * PublicRecordsLast10Years
    * LoanOriginalAmount

####  What is/are the main feature(s) of interest in your dataset?

The main features of interest in this dataset are `BorrowerRate` and `ProsperRating`, where `ProsperRating` might be the overall assessment and main quality indicators of borrowers performed by Prosper. Also, I've seen evidence that `LenderYield` is the most important factor for investors.

####  What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

Features om the dataset that I think will help support my investigation into features of interest are variables regarding borrower's status (i.e.`EmploymentStatus`, `IsBorrowerHomeowner`, etc.) as well as those that reveal borrower's past payment history (i.e. `PublicRecords`, `Inquiries`, etc.). Also, I did not find anything unusual about `IncomeRange` and `IncomeVerifiable`, variables that I was expecting more from. On the other hand, maybe this variable have more to tell when related to others.

####  Did you create any new variables from existing variables in the dataset?

I factorized a few variables because those variables are either intrinsically categorical or they only have several possible values.

#### Of the features you investigated, were there any unusual distributions? 
#### Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

Of the features you investigated, I observed some unusual distributions in the `LoanOriginalAmount` histogram. The frequencies are higher around `4,000`, `5,000`, `10,000`, `15,000` and so on. I believe these amounts are more common for most borrowers to choose from.

I did not perform any operations on the dataset, as the data was tidy, Also, I did not perform other transformation other than changing their classes (e.g. numeric -> factor ) since categorical variables enter into statistical models differently than continuous variables. Thus, storing data as factors ensures that the modeling functions will treat such data accordingly.

------

# Bivariate Plots Section

```{r echo=FALSE, fig.height=8, warning=FALSE, Correlation }
# Correlation matrix of numerical variables
# Text and number sizes are adjusted
ggcorr(
  data = loandata,
  label = TRUE,
  label_size = 1.8,
  size = 3,
  hjust = 0.71
) +
  theme_pubr(margin = FALSE,
             border = FALSE,
             legend = 'right')
```

> The correlation matrix shows that there are two strong positive relationships among the features.
The trivial features are `CreditScoreRangeUpper` and `CreditScoreRangeLower`. Next are the `BorrowerRate` and `LenderYield.`

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, BorrowerRateLenderYield }
# Scatter plot: BorrowerRate and Lender Yield
ggplot(data = loandata, aes(x = BorrowerRate, y = LenderYield)) +
  geom_point(fill = I('#96ADBD'), color = I('black'), shape = 21) +
  geom_smooth(method = "lm", color = I('red'))  + # Add regression line 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
  theme_pubr()

# Calcuate the correlation coefficient
cor(loandata$BorrowerRate, loandata$LenderYield)
```
> The `BorrowerRate` and the `LenderYield` have a near perfect linear relationship. This is a result from Prosper charging a fixed rate on loans. Therefore, the `BorrowerRate` and the `LenderYield` should have a strong positive relationship.

```{r echo=FALSE, fig.height=8, message=FALSE, warning=FALSE, Rating_vs_BorrowerRate }
# Boxplot: ProsperRating and BorrowerRate
ggplot(data = subset(loandata, ProsperRating != ""),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  labs(caption = "*NA excluded") +
  ggtitle("Borrower Rate by Prosper Rating") +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5)) 
```

> This boxplot shows that `ProsperRating` does affect the `BorrowerRate` and its relationship.

Now, I want to inspect other factors that also have impact on the `BorrowerRate` and the `ProsperRating`. Let's look at the `BorrowerRate` next.

```{r, echo=FALSE, message=FALSE, warning=FALSE, BorrowerRate_vs_Term }
# Boxplot: Term and BorrowerRate
ggplot(data = loandata, aes(x = Term, y = BorrowerRate)) +
  geom_boxplot(color = I('#425E72')) +
  ggtitle("Borrower Rate by Term") +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5)) 

# Summaries of BorrowerRate by Term
tapply(loandata$BorrowerRate, loandata$Term, summary)
```
> The average rate of the 3-year term is larger than the 1-year term. However, these terms are very close to the 5-year term average. Also, loans with a 5-year term average have a smaller range than the 3-year term group.

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, OrignalAmount_vs_BorrowerRate}
# Scatter plot: LoanOriginalAmount and BorrowerRate
ggplot(data = loandata, aes(x = LoanOriginalAmount, y = BorrowerRate)) +
  geom_point(alpha = 0.2, color = '#425E72') +
  # geom_point(alpha = 0.2, fill = '#96ADBD', color = 'black', shape = 21) +  
  ggtitle("Borrower Rate by LoanOriginalAmount") +
  theme_pubr() +
  theme(plot.title = element_text(hjust = 0.5)) 

# Calcuate the correlation coefficient
cor(loandata$LoanOriginalAmount, loandata$BorrowerRate)
```
> This graph shows that there is a fairly weak negative relationship between `BorrowerRate` and `OriginalLoanAmount`.

```{r, echo=FALSE, fig.height=8, message=FALSE, warning=FALSE, BorrowerRate_vs_ListingCategory}
# Boxplot: ListingCategory and BorrowerRate
ggplot(data = loandata, aes(x = ListingCategory, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  coord_flip() +
  theme_pubr()
```

> It seems that some categories have lower rates and others have higher rates.

```{r echo=FALSE, IncomeVerificable_and_Homeowner }
# Boxplot: IncomeVerifiable and BorrowerRate
ggplot(data = loandata, aes(x = IncomeVerifiable, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  theme_pubr()

# Boxplot: IsBorrowerHomeowner and BorrowerRate
ggplot(data = loandata, aes(x = IsBorrowerHomeowner, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  theme_pubr()
```

> Borrowers that are income verifiable or homeowners seem to have lower rates, which makes sense.

Finally, let's take a look at how `ProsperRating` affect the aforementioned features.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_Ratings}
# Histogram of Prosper Rating by numbers of loans
ggplot(aes(ProsperRating), data = loandata) +
  geom_histogram(stat = 'count', aes(fill = ProsperRating)) +
  ggtitle('Number of Loans by Prosper Rating') +
  labs(x = 'Rating', y = 'Loan Count') +
  theme_pubr(legend = 'right') +
  theme(plot.title = element_text(hjust = 0.5))  

summary(loandata$ProsperRating)
```
> The majority of loans in this dataset are non-classified (i.e.'NA'), but, among those that are classified, loans with a 'C' rating is the highest.

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, ProsperRating_vs_CreditScore}
# Boxplot: ProsperRating and CreditScoreRangeUpper
ggplot(data = subset(loandata, ProsperRating != ""),
       aes(x = ProsperRating, y = CreditScoreRangeUpper)) +
  geom_boxplot(color = '#425E72') +
  labs(caption = "*NA excluded") +
  theme_pubr() +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

> A higher credit score by the consumer rating agency also tends to have a higher `ProsperRating`.

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, ProsperRating_vs_IsBorrowerHomeowner }
# Bar charts: ProsperRating colored by IsBorrowerHomeowner
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, fill = IsBorrowerHomeowner)) +
  geom_bar(position = "dodge") +
  scale_fill_discrete() +
  scale_fill_manual(values = c("#96ADBD", "#425E72")) +
  labs(x = 'Rating', y = 'Loan Count', caption = "*NA excluded") +
  theme_pubr(legend = 'right') +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> For top `ProsperRating` groups like "AA" and "A", Homeowners are the majority.

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, ProsperRating_vs_IncomeVerifiable }
# Bar charts: ProsperRating colored by IncomeVerifiable
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, fill = IncomeVerifiable)) +
  geom_bar(position = "dodge") +
  scale_fill_discrete() +
  scale_fill_manual(values = c('#96ADBD', '#425E72')) +
  labs(x = 'Rating', y = 'Loan Count', caption = "*NA excluded") +
  theme_pubr(legend = 'right') +  
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> This chart shows evidence that it is difficult to get a good rating if a borrower's income not verifiable

## Bivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
First, the `LenderYield` and `BorrowerRate` are strongly related because the fixed rate charged by Prosper. Second, the `BorrowerRate` and `ProsperRating` do vary with borrower status. In addition, a good relationships I found was through the proprietary Prosper Rating system.

#### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
The relationship between `ListCategory` and `BorrowerRate` is interesting. In average, for this relationship, cosmetic procedure's rates are relatively higher, whereas boat rates are lower. 

#### What was the strongest relationship you found?
The strongest relationship I found is the one between the `LenderYield` and `BorrowerRate`, with a correlation coefficient very close to 1. In addition, the relationship between `ProsperRating` / `LenderYield` and `ProsperRating` / `BorrowerRate` have an inverse relationship. The higher the rating, the lower the `LenderYield` and `BorrowerRate`.

------

# Multivariate Plots Section

Now let's put everything together based on what I found in the bivariate plots section by creating a few multivariate plots to investigate more complex interactions between variables. 

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, fig.height=7, Jitter_Plot_Homeowner }
# Jitter plot: ProsperRating and BorrowerRate, colored by IsBorrowerHomeowner
ggplot(data = subset(loandata, loandata$ProsperRating != ''),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_jitter(aes(color = IsBorrowerHomeowner), alpha = 0.4) +
  scale_color_brewer(type = 'qual', palette = 'Set1') +
  labs(caption = "*NA excluded") +
  theme_pubr(legend = 'right')  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> Using jitter plot, we can visualize the real distribution of `BorrrowerRate` relative to `ProsperRating`. In addition, I use different colors to indicate whether the individual borrower is a homeowner. This plot shows that most of the "AA" rated borrowers are also homeowners.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, Colored_Scatter_Plot_OriginalAmount }
# Scatter plot: LoanOriginalAmount and BorrowerRate, colored by ProsperRating
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = LoanOriginalAmount, y = BorrowerRate, color = ProsperRating)) +
  geom_point(alpha = 0.5) +
  scale_color_brewer(type = "qual", palette = "Paired") +
  labs(caption = "*NA excluded") +
  theme_pubr(legend = 'right')  
```

> In this plot, I found evidence that each rating has its own colored area that is easily recognized, which indicates that `ProsperRating` is a good predictor for the `BorrowerRate`. Also, highly rated loans (i.e. "AA", "A", and "B") can have loan original amount over $30,000. On the other hand, most lowly rated loans (i.e. "E" and "HR") are below $10,000.

```{r echo=FALSE, message=FALSE, warning=FALSE, Heat_Map_ListCategory_summarize }
# Create the frequency table
subset_table <- subset(loandata, loandata$ProsperRating != "") %>%
  dplyr::select(ProsperRating, ListingCategory, BorrowerRate) %>%
  dplyr::group_by(ProsperRating, ListingCategory) %>%
  summarize(BorrowerRate = mean(BorrowerRate))

head(subset_table)
```

```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, Heat_Map_ListCategory }
# Heat map: BorrowerRate based on ProsperRating nad ListingCategory 
ggplot(aes(x = ProsperRating, y = ListingCategory
           , fill = BorrowerRate), data = subset_table) +
  geom_tile(stat = "identity") +
  scale_fill_gradientn(colours = colorRampPalette(c("lightblue", "navy"))(100)) +
  labs(caption = "*NA excluded") +
  theme_pubr(legend = 'right')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

> This plot shows the relative value of the average `BorrowerRates` of different `ListCategory`. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=7, Facet_Boxplot_Category }
# Faceted boxplot: ProsperRating and BorrowerRate by ListingCategory
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  facet_wrap( ~ ListingCategory) +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> Compared to the previous plot, the boxplots give us more information about the outliers and range of the distribution. However, it's more difficult to compare mean values between different `Listingcategory`.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, Jitter_Plot_Facet_ListCategory }
# Faceted jitter plots: ProsperRating and BorrowerRate by ListingCategory
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_jitter(alpha = 0.2, color = '#425E72') +
  facet_wrap(~ ListingCategory) +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

> Based on this plot, 'Boat', 'Green loans', and 'RV' categories have less data points. As a result, these catefories have less outliers as seen in the previous plots.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height = 7, Three_Plots_EmploymentStatus_summary }
# Create the frequency table
table <- subset(loandata, loandata$ProsperRating != "") %>%
    dplyr::select(ProsperRating, EmploymentStatus, BorrowerRate) %>%
    dplyr::group_by(ProsperRating, EmploymentStatus) %>%
    summarize(BorrowerRate = mean(BorrowerRate))

# Heat map: BorrowerRate based on ProsperRating nad EmploymentStatus
ggplot(data = table,
       aes(x = ProsperRating,
           y = EmploymentStatus,
           fill = BorrowerRate)) +
  geom_tile(stat = "identity")+
  scale_fill_gradientn(colours = colorRampPalette(c("lightblue", "navy"))(100)) +
  labs(caption = "*NA excluded") +
  theme_pubr(legend = 'right')+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Faceted boxplot: ProsperRating and BorrowerRate by EmploymentStatus
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_boxplot(color = '#425E72') +
  facet_wrap( ~ EmploymentStatus, nrow = 4) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  

# Faceted jitter plots: ProsperRating and BorrowerRate by EmploymentStatus
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
       aes(x = ProsperRating, y = BorrowerRate)) +
  geom_jitter(alpha = 0.2, color = '#425E72') +
  facet_wrap( ~ EmploymentStatus, ncol = 2) +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> The heat map shows that highly rated "Not employed" borrowers have to pay slight higher rates. The jitter plots shows that even "Not employed", "Retired", or "Part-time" borrowers can get loan with rates lower than 0.2.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=8, Jitter_Plots_Term }
# Faceted jitter plots: ProsperRating and BorrowerRate by Term
plot1 <- ggplot(data = subset(loandata, loandata$ProsperRating != ""),
                aes(x = ProsperRating, y = BorrowerRate, color = ProsperRating)) +
  geom_point(alpha = 0.1, position = 'jitter') +
  facet_grid(. ~ Term) +
  labs(caption = "*NA excluded") +
  ggtitle("Borrower Rate vs Term and Prosper Rating") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot2 <- ggplot(data = subset(loandata, loandata$ProsperRating != ""),
                aes(x = ProsperRating, y = BorrowerRate)) +
  geom_boxplot(aes(fill = ProsperRating)) +
  facet_grid(. ~ Term) +
  labs(caption = "*NA excluded") +
  ggtitle("Boxplot - Borrower Rate vs Term and Prosper Rating") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(plot1, plot2, ncol=1, nrow =2)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Jitter_Plots_Term_Calcs_Printed }
# Calculate means for each Term
print(kable(
  tapply(loandata$BorrowerRate, loandata$Term, mean),
  col.names = "mean",
  align = 'lc'))

# Calculate standard deviations for each Term
print(kable(
  tapply(loandata$BorrowerRate, loandata$Term, sd),
  col.names = "standard deviation",
  align = 'lc'))
```
> This chart shows a closer look at `BorrowerRate` and `ProsperRating` in which  most loans have either a 3-year or 5-year term with higher borrower rates. On the other hand, the main difference between 3-year and 5-year term loans is deviation.

```{r echo=FALSE, fig.width = 12, fig.height = 10, BorrowerRate_Year_Term}
# Borrower Rate vs Year vs Prosper Rating
# set Loan Orignal Date in Year-Month-Date format
loandata$LoanOriginationDate <-
  as.POSIXct(loandata$LoanOriginationDate,
             format = "%Y-%m-%d")
loandata$LoanOriginationDate.year <-
  loandata$LoanOriginationDate %>%
  format('%Y') %>% strtoi()

plot3 <- ggplot(aes(x = ProsperRating, y = BorrowerRate),
                data = na.omit(filter(loandata))) +
  geom_boxplot(aes(fill = ProsperRating)) +
  facet_grid(. ~ LoanOriginationDate.year) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Borrower Rate vs Term and Prosper Rate")

plot4 <-
  ggplot(aes(x = ProsperRating, y = BorrowerRate
             , color = ProsperRating),
         data = na.omit(filter(loandata))) +
  geom_point(alpha = 1, position = "jitter") +
  facet_grid(. ~ LoanOriginationDate.year) +
  ggtitle("Borrower Rate vs Term and Prosper Rate") +
  labs(caption = "*NA excluded") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

grid.arrange(plot3, plot4, ncol=1, nrow =2)
```

> There's evidence that Prosper optimized their rating model throughout the year, as we see the borrower and the variation between borrower rate is not that significant anymore. As a result, we tend to have smaller standard deviation year-over-year. Something to note is the amount of borrowing that suddenly decreased in 2013.

## Multivariate Analysis

#### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
A feature that strengthened each other in terms of looking at the features of interest is `Term`, as this is a good indicator whether we have a better `BorrowerRate` or not. In addition, we see how variables like `BorrowerRate` and `ProsperRating` come together and how it affect each other. 

#### Were there any interesting or surprising interactions between features?
The criteria for being an AA borrowers seems to be tighten year-over-year. Also, there seems to be a fixed borrower rate for both the HR and AA criteria.

#### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.
No, I did not create any models with this dataset.

------

# Final Plots and Summary

At this point, I've done some exploration and have built up an understanding of the structure of and relationships between the variables in this Prosper Loan dataset. Next, I will select three plots from all of my previous exploration to present here as a summary of some of my most interesting findings.

### Plot One
```{r echo=FALSE, fig.height=8, message=FALSE, warning=FALSE, warning=FALSE, Plot_One}
# Subplot 1: Boxplot
p3 <- ggplot(data = subset(loandata, ProsperRating != ""),
             aes(x = ProsperRating, y = BorrowerRate * 100)) +
  geom_boxplot(color = '#425E72') +
  labs(
    title = "Boxplot and Jitter Plots",
    subtitle = "Borrower rate by Prosper rating",
    x = "",
    y = "Borrower Rate (%)") +
  theme_pubr() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)
        , plot.subtitle = element_text(size = 10, color = "#425E72"))  

# Subplot 2: Jitter plot
p4 <- ggplot(data = subset(loandata, loandata$ProsperRating != ""),
             aes(x = ProsperRating, y = BorrowerRate * 100)) +
  geom_jitter(alpha = 0.2, color = '#425E72') +
  labs(x = "Prosper Rating",
       y = "Borrower Rate (%)") +
  theme_pubr() +
  labs(caption = "*NA excluded") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)
        , plot.subtitle = element_text(size = 10, color = "#425E72"),
        plot.caption = element_text(color = "#425E72", face = "italic"))  

# Combine above two plots
grid.arrange(p3, p4, ncol = 1)
```

### Description One
The Prosper rating is an important factor for the borrower rate. In other words, a high Prosper rating is linked to lower borrower rates and vice versa.

### Plot Two
```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, Plot_Two}
# Colored scatter plot: LoanOriginalAmount, BorrowerRate, ProsperRating
ggplot(data = subset(loandata, loandata$ProsperRating != ""),
  aes(x = LoanOriginalAmount, y = BorrowerRate * 100, color = ProsperRating)) +
  geom_point(alpha = 0.5) +
  scale_color_brewer(type = "qual", palette = "Set1") +
  labs(
    title = "Borrower Rate by Prosper Rating and Loan Original Amount",
    x = "Loan Original Amount($)",
    y = "Borrower Rate (%)",
    color = "Prosper Rating") + 
  theme_pubr(legend = 'right') +
  theme(legend.background = element_rect(color = 'black'),
        plot.title = element_text(hjust = 0.5, vjust = 1, size = 14, face = "bold")) 
```

### Description Two
Based on this graph, highly rated loans (i.e. "AA", "A", and "B") can have original amount over $30,000, and most lowly rated loans (i.e. "E" and "HR") are under 10,000.  As a results, Prosper rating is an great indicator of the borrower rate. As you can see, the different colored areas can be easily recognized.

### Plot Three
```{r echo=FALSE, fig.height=7, message=FALSE, warning=FALSE, Plot_Three}
p5 <- ggplot(data = subset(loandata, loandata$ProsperRating != ""),
             aes(x = ProsperRating, y = BorrowerRate * 100, color = ProsperRating)) +
  geom_point(alpha = 0.1, position = "jitter") +
  facet_grid( ~ Term) +
  labs(title = "Borrower rate by Prosper rating and term",
       x = "Prosper Rating",
       y = "Borrower Rate (%)",
       caption = "*NA Excluded",
       color = "Prosper Rating") +
  theme_pubr(legend = 'right') +
  scale_color_brewer(type = 'qual', palette = 'Set1') +  
  theme(legend.background = element_rect(color = 'black'),
        plot.title = element_text(hjust = 0.5, vjust = 1, size = 14, face = "bold"),
        legend.key = element_rect(fill = "lightgrey", colour = "black"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.caption = element_text(color = "red", face = "italic")) 

# Add means and stds to the plot
ggadd(p5, c("mean_sd"), color = "black")
```

### Description Three
This graph shows eveidence that a 3-year term loan will have higher borrower rates when compared to a 1-year term loan in each Prosper rating. In addition, loans with a 5-year term are more concentrated than a 3-year term loan and has a smaller total range as well.

------

# Reflection

Regarding my initial observations on the Prosper loan dataset containing up to 81 variables, I was considering changing datasets because I believed the high variable count would make it hard to explore and analyze. However, after examining the dataset's variable dictionary that explained the variables in the data set, I was able to determine which variables might be interesting to explore. Furthermore, I researched and explored each variable primarily focusing on features related to the borrower rates and statuses. Next, before diving into my exploratory data analysis, I took a subset of the loan data, creating a smaller dataset in which reduced the variable count to 15.

In my updated dataset, there were only 7 numerical variables. Initially, my primary focus was to start analyzing the borrower rate and the various borrower statuses. However, I decided to go through each variable one by one, trying to understand any relationships between the variables in the dataset since I'm currently working in the financial industry. As a result, most of the data visualizations I created included at least one categorical variable. 

I did spend a lot of time researching how to create visualization with one, two, or more categorical variables. One of my main struggles, which is not uncommon, was the material that I've learned up to this point were outdated, so I ended up doing additional research on the latest and up-to-date libraries and functions needed to complete this project. Based on my research and what I've learned, my solution was to utilize boxplots, bar charts, jitter plots, heat maps, and a variety of libraries/functions to create the visualizations needed for this project. 

In conclusion of my project, my analysis shows that Prosper rating and term are two major indicators of the borrower rate. In addition, borrower statuses also have impacts on borrower rate, but their effects are either small or are also factors of Prosper rating. Overall, I think this project would be a good reference for anyone who wants to borrower money from a P2P lending platform. For future work and to expand the project, I can try to combine various variables and make a predictive model using logistic regression for a go/no-go investment decision or any classification algorithm would make sense. Overall, I enjoyed this project. I will say that my exploratory data analysis did give me better insight into the lending business.